Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-06-05 TEX/draft4.tex
Still working on discussion, and resolved some of the questions in footnote by chatting with tamir

\TODO{Revise--Cryo--electron microscopy (cryo--EM) is an innovative technology for single particle reconstruction (SPR) of macromolecules.} 
%In recent years, structures of many molecules, previously regarded as insurmountable, are now being reconstructed to near-atomic resolution
%; see for instance~\cite{kuhlbrandt2014resolution,bartesaghi20152,smith2014beyond}.
%This technological advancement was recognized by the 2017 Nobel Prize in Chemistry~\cite{nobel}. 
% 
In a cryo--EM experiment, biological samples are rapidly frozen in a thin layer of vitreous ice. Within the ice, the molecules are randomly oriented and positioned. The microscope produces a 2D image of the samples embedded in the ice called a \emph{micrograph}. Each micrograph contains tomographic projections of the samples at unknown locations and under unknown viewing directions. The goal is to construct 3-D models of the molecules from the micrographs.

@@

%Besides the model bias, particle pickers have several limitations due to the high noise level, in particular for small particles and in low contrast. 
%As a result, the projections in the 2D images are typically not centered; including the correct centers as parameters dramatically increases the complexity of reconstruction algorithms.
%In addition, the information from particles that are too close to each other is usually neglected. Hence, valuable information that can be harnessed is omitted.

@@

%To support the claim that it may be possible to bypass particle picking in SPR from micrographs, as a proof of concept,
We conducted two experiments in the simplified image formation model described in the introduction:
\begin{enumerate}
	\item The first experiment aims to recover a 2D image from an increasing number of micrographs with high noise, similar to the rightmost panel of Figure~\ref{fig:micro_example}. This is done using moments of \TODO{first and} second order, as these are sufficient to recover a 2D image up to elementary symmetries;
	\item The second experiment aims to recover three distinct 1D signals from an increasing number of 1D micrographs with high noise. For this task, it is necessary to use moments up to third order.
\end{enumerate}
As outlined below, we find that it is indeed possible to recover good estimates of the ground truth signals from the highly corrupted micrographs, without particle picking. Furthermore, we find that the quality of estimation increases with the amount of data collected, despite the fact that particle picking remains challenging. The methods section provides additional details. In the discussion section, we outline how the general approach could be extended to full 3D SPR.

@@

In the first experiment, we estimated Einstein's image of size $50\times 50$ from a growing number of micrographs, each of size $4096\times 4096$ pixels. A micrograph contains, on average, 700 occurrences of the target image at random locations. The latter are chosen so that two occurrences are always separated by at least 49 pixels. Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$ (this corresponds to $\SNR=1/20$). This high noise level is illustrated in the right panel of Figure~\ref{fig:micro_example}.

We compute the average autocorrelation of the micrographs (equivalently, the average of their power spectra). This is a particularly simple computation. In the methods section, we show how, owing to separation of the occurrences, a determined portion of the averaged autocorrelation allows to estimate the power spectrum of the unknown image itself. Mathematically, it is easy to show that the quality of this estimate improves steadily as the amount of data grows, regardless of noise level. Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{bauschke2004rrr,elser2017rrr}.\footnote{\TODO{What about biased terms? Do we need to know $\sigma$ in the 2D experiment? What about the number of occurrences, do we need to know that? Quid of symmetries? And quid first order moments?}} RRR is initialized with a picture of the physicist Isaac Newton and iterates to produce the estimate.\footnote{\TODO{Initializing with a face is debatable: it might helps the algorithm. On the other hand, I agree that it supports the notion that we don't get trapped too easily.}}

@@

%\begin{figure}[h!]
%	\centering
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{Newton}
%		\caption{\label{fig:newton}Newton  (model)}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction2}
%		\caption{\small $P =1024$}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction20}
%		\caption{$P = 1024\times 10$}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction200}
%		\caption{$P = 1024\times 100$}
%	\end{subfigure}%
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized with an image of the physicist Isaac Newton. Estimates are obtained from $P$ micrographs (growing across panels), each containing $700$ image occurrences on average.---\TODO{We may want to add another image with $P=512\times 10^4$ and the image of Newton.}
		% At a noise level of $\sigma = 3$, this amounts to an $\SNR$ of $1/20$.
	}
\end{figure}

@@

In the second experiment, three 1D signals, each of length $L = 21$, appear at random locations in one long 1D signal, which we call a micrograph by analogy. Any two occurrences are separated by at least 20 entries. The signals appear respectively about 30, 20 and 10 million times in a micrograph of length 12.3 billion. The micrograph is then contaminated with additive white Gaussian noise. This results in an $\SNR$ of about $1/9$, while about 10\% of the micrograph contains signal. Neither the number of occurrences nor the noise level $\sigma$ are known.

@@

\section{Discussion}

\TODO{Discussion: What might the answer imply and why does it matter? How does it fit in with what other researchers have found? What are the perspectives for future research?}

@@

In this paper, we examined a model that captures certain important features of the SPR problem. In this model, the aim is to estimate one, or possibly several, images from micrographs. To this end, we compute the autocorrelation functions of the data and relate them to the unknown parameters. This leads to a system of polynomial equations which must be solved.

The same general approach can be applied directly to SPR from cryo-EM. In this scenario, the micrographs contain numerous tomographic projections of molecules (possibly in different conformations) taken under unknown viewing directions. The aim here is to estimate the 3-D volumes of the different conformations directly from micrographs. In Appendix~\TODO{ref}, we show that the number of polynomial equations provided by the third-order autocorrelations is of the same order as the number of unknown volume coefficients. This hints that it may be possible to reconstruct a constant number of distinct volumes from these equations directly. Crucially, this involves no particle picking, and hence also does not involve a viewing direction estimation stage for individual particles. \TODO{Where and how do we cite Kam? Fred?}

@@

As stated, our approach relies on two core assumptions that are not necessarily verified in SPR experiments.
First, we assume an additive white noise model, while in practice the noise may be structured or signal dependent. To address this point, it may be necessary to investigate better noise models and to extend the autocorrelation analysis to these models.
Second, we assume that any two signal occurrences are sufficiently separated, as required by~\eqref{eq:spacing}. 
This separation can be induced by careful experimental design \TODO{Ref?}.
If the signals are not well separated, one can introduce new variables that represent the distribution of the spacing between signal occurrences. Here as well, the relation between the autocorrelation functions of the data and the signals can be derived. These variations lead to different inverse problems. No sufficient conditions for them to be solvable are currently known.

@@

%For the 1D experiment, we fix $K = 3$ signals of length $L = 21$, as depicted in Figure~\TODO{ref}. Following the data model described in Section~\TODO{ref}, we generate an observation $y$ of length $24.6 \cdot 10^9$. Each of the three signals appears, respectively (and approximately) $300 \cdot 10^6$, $200 \cdot 10^6$ and $100 \cdot 10^6$ times in $y$, such that at least $L-1$ zeros separate two occurrences of any signals. 
%This is done by randomly selecting $600 \cdot 10^6$ placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far. For each placement, one of the three signals is picked at random proportionally to the desired number of occurrences of each. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The SNR of $y$
% sqrt((m_want*sum(X.^2)')/(sigma^2*n))
%is about 1/12.

@@

\section{The autocorrelation functions in cryo--EM}

>>>>

\TODO{Revise--Cryo--electron microscopy (cryo--EM) is an innovative technology for single particle reconstruction (SPR) of macromolecules.} 
%In recent years, structures of many molecules, previously regarded as insurmountable, are now being reconstructed to near-atomic resolution
%; see for instance~\cite{kuhlbrandt2014resolution,bartesaghi20152,smith2014beyond}.
%This technological advancement was recognized by the 2017 Nobel Prize in Chemistry~\cite{nobel}. 
% 
In a cryo--EM experiment, biological samples are rapidly frozen in a thin layer of vitreous ice. Within the ice, the molecules are randomly oriented and positioned. The microscope produces a 2-D image of the samples embedded in the ice called a \emph{micrograph}. Each micrograph contains tomographic projections of the samples at unknown locations and under unknown viewing directions. The goal is to construct 3-D models of the molecules from the micrographs.

@@

%Besides the model bias, particle pickers have several limitations due to the high noise level, in particular for small particles and in low contrast. 
%As a result, the projections in the 2-D images are typically not centered; including the correct centers as parameters dramatically increases the complexity of reconstruction algorithms.
%In addition, the information from particles that are too close to each other is usually neglected. Hence, valuable information that can be harnessed is omitted.

@@

%To support the claim that it may be possible to bypass particle picking in SPR from micrographs, as a proof of concept,
We conducted two experiments in the simplified image formation model described in the introduction:
\begin{enumerate}
	\item The first experiment aims to recover a 2-D image from an increasing number of micrographs with high noise, similar to the rightmost panel of Figure~\ref{fig:micro_example}. This is done using moments of second order, as these are sufficient to recover a 2-D image up to elementary symmetries;
	\item The second experiment aims to recover three distinct 1-D signals from an increasing number of 1-D micrographs with high noise. For this task, it is necessary to use moments up to third order.
\end{enumerate}
As outlined below, we find that it is indeed possible to recover good estimates of the ground truth signals from the highly corrupted micrographs, without particle picking. Furthermore, we find that the quality of estimation increases with the amount of data collected, despite the fact that particle picking remains challenging. The methods section provides additional details. In the discussion section, we outline how the general approach could be extended to full 3D SPR.

@@

In the first experiment, we estimated Einstein's image of size $50\times 50$ and mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. A micrograph contains, on average, 700 occurrences of the target image at random locations. The latter are chosen so that two occurrences are always separated by at least 49 pixels. Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$ (this corresponds to $\SNR=1/20$). This high noise level is illustrated in the right panel of Figure~\ref{fig:micro_example}. In this first experiment, we assume knowledge of $\sigma$ and of the total number of signal occurrences across all micrographs.

We compute the average autocorrelation of the micrographs (equivalently, the average of their power spectra). This is a particularly simple computation. In the methods section, we show how, owing to separation of the occurrences, a determined portion of the averaged autocorrelation allows to estimate the power spectrum of the unknown image itself. Mathematically, it is easy to show that the quality of this estimate improves steadily as the amount of data grows, regardless of noise level. Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{bauschke2004rrr,elser2017rrr}.\footnote{\TODO{Quid of symmetries?}} RRR is initialized with a picture of the physicist Isaac Newton and iterates to produce the estimate.\footnote{\TODO{Initializing with a face is debatable: it might help the algorithm. On the other hand, I agree that it supports the notion that we don't get trapped too easily.}}

@@

%\begin{figure}[h!]
%	\centering
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{Newton}
%		\caption{\label{fig:newton}Newton  (model)}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction2}
%		\caption{\small $P =1024$}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction20}
%		\caption{$P = 1024\times 10$}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction200}
%		\caption{$P = 1024\times 100$}
%	\end{subfigure}%
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized with an image of the physicist Isaac Newton. Estimates are obtained from $P$ micrographs (growing across panels), each containing $700$ image occurrences on average.---\TODO{Will add image with $P=512\times 10^4$ and image of Newton.}
		% At a noise level of $\sigma = 3$, this amounts to an $\SNR$ of $1/20$.
	}
\end{figure}

@@

In the second experiment, three 1-D signals, each of length $L = 21$, appear at random locations in one long 1-D signal, which we call a micrograph by analogy. Any two occurrences are separated by at least 20 entries. The signals appear respectively about 30, 20 and 10 million times in a micrograph of length 12.3 billion. The micrograph is then contaminated with additive white Gaussian noise. This results in an $\SNR$ of about $1/9$, while about 10\% of the micrograph contains signal. Neither the number of occurrences nor the noise level $\sigma$ are known.

@@

\clearpage
\section{Discussion}

%\TODO{Discussion: What might the answer imply and why does it matter? How does it fit in with what other researchers have found? What are the perspectives for future research?}

@@

\TODO{Where and how do we cite Kam? Fred?}

In the simplified model we examined, the aim is to estimate one, or possibly several, images from micrographs. Our strategy is to compute autocorrelation functions of the data and to relate these statistics to the unknown parameters. Recovering the parameters from the statistics reduces to solving a set of polynomial equations. Depending on the scenario, we did so using either a phase retrieval algorithm or a nonlinear least-squares algorithm.

The same general approach can, in principle, be applied directly to SPR from cryo-EM. Here, the micrographs contain numerous tomographic projections of molecules (possibly in different conformations) under unknown viewing directions. The aim is to estimate the 3-D volumes of the different conformations directly from micrographs. Each volume can be expanded linearly in a basis, so that the volume is characterized by its expansion coefficients. Since tomographic projection is a linear operation, autocorrelations of the micrographs (which can be estimated easily) are polynomial functions of the sought coefficients. Thus, autocorrelations of the micrographs provide a system of polynomial equations in the volume parameters, and the question becomes: are these equations sufficient to uniquely identify the volumes, and can we solve the system?

We show in Appendix~\ref{sec:cryoem} that the number of polynomial equations provided by the third-order autocorrelations is of the same order as the number of coefficients required to describe one volume at the same resolution as the particle projections. This hints that it may be possible to reconstruct a constant number of distinct volumes from these equations directly. Crucially, the outlined approach involves no particle picking, and a fortiori no viewing direction estimation. As a result, it may not be limited to large molecules in the same way that particle picking approaches are. Concerns for model bias would also be greatly reduced.

@@

As stated, our approach relies on two core assumptions that are not necessarily verified in SPR experiments.
First, we assume an additive white noise model, while in practice the noise may be structured or signal dependent. To address this point, it may be necessary to investigate better noise models and to extend the autocorrelation analysis to these models.
Second, we assume that any two signal occurrences are sufficiently separated, as required by~\eqref{eq:spacing}. 
This separation could be induced by careful experimental design \TODO{Ref?}.
If the signals are not well separated, one can introduce new variables that represent the distribution of the spacing between signal occurrences. Here as well, the relation between the autocorrelation functions of the data and the signals can be derived. These variations lead to different inverse problems. No sufficient conditions for them to be solvable are currently known.

@@

%For the 1-D experiment, we fix $K = 3$ signals of length $L = 21$, as depicted in Figure~\TODO{ref}. Following the data model described in Section~\TODO{ref}, we generate an observation $y$ of length $24.6 \cdot 10^9$. Each of the three signals appears, respectively (and approximately) $300 \cdot 10^6$, $200 \cdot 10^6$ and $100 \cdot 10^6$ times in $y$, such that at least $L-1$ zeros separate two occurrences of any signals. 
%This is done by randomly selecting $600 \cdot 10^6$ placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far. For each placement, one of the three signals is picked at random proportionally to the desired number of occurrences of each. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The SNR of $y$
% sqrt((m_want*sum(X.^2)')/(sigma^2*n))
%is about 1/12.

@@

\section{The autocorrelation functions in cryo--EM} \label{sec:cryoem}
